Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-01-17 cordillera-cycle.tex
Typeset: whitespace changes in manuscript.
Also one found one typo: aroung -> around.

      In general, the topographic complexity of the North American
      Cordillera and its effect on glacial history have inhibited the
      reconstruction of ice sheet-wide glacial advance and retreat patterns
      such as those available for the Fennoscandian and Laurentide ice
      sheets \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
      Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}. Here, we
      use a~numerical ice sheet model \citep{PISM-authors.2015}, calibrated
      against field-based evidence, to perform a~quantitative reconstruction
      of the Cordilleran ice sheet evolution through the last glacial cycle,
      and analyse some of the long-standing questions related to its
      evolution:

\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the LGM?
  \item What was the scale of glaciation prior to the LGM?
  \item Which were the primary dispersal centres? Do they reflect stable or
    ephemeral configurations?
  \item How rapid was the last deglaciation? Did it include Late Glacial
    standstills or readvances?
\end{itemize}

      Although numerical ice sheet modelling has been established as
      a~useful tool to improve our understanding of the Cordilleran ice sheet
      (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
      \citealp{Marshall.etal.2000}), the ubiquitously mountainous topography
      of the region has presented two major challenges to its application.
      First, only recent developments in numerical ice sheet models and
      underlying scientific computing tools \citep{Bueler.Brown.2009,
      Balay.etal.2015} have allowed for high-resolution numerical modelling
      of glaciers and ice sheets on mountainous terrain over millennial time
      scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the complex
      topography of the North American Cordillera also induces strong
      geographic variations in temperature and precipitation, thus requiring
      the use of high-resolution climate forcing fields as an input to an
      ice sheet model \citep{Seguinot.etal.2014}. However, evolving climate
      conditions over the last glacial cycle are subject to considerable
      uncertainty and still lie beyond the computational reach of atmosphere
      circulation models.

@@

      The PDD computation accounts for stochastic temperature variations by
      assuming a~normal temperature distribution of standard deviation
      $\sigma$ aroung the expected value $T_{\mathrm{m}}$. It is expressed by an
      error-function formulation \citep{Calov.Greve.2005},
\begin{align}
&\label{eqn:calovgreve}
    {\text{PDD}} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{\mathrm{m}}^2}{2\sigma^2}}\right)
              + \frac{T_{\mathrm{m}}}{2} \, {\text{erfc}}
                \left(-\frac{T_{\mathrm{m}}}{\sqrt{2}\sigma}\right)\right] \,,
\end{align}
      which is numerically approximated using week-long sub-intervals. In
      order to account for the effects of spatial and seasonal variations of
      temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
      from NARR daily temperature values from 1979 to 2000
      \citep{Mesinger.etal.2006}, including variability associated with the
      seasonal cycle (Fig.~\ref{fig:atm}). Degree-day factors for snow and
      ice melt are derived from mass-balance measurements on contemporary
      glaciers from the Coast Mountains and Rocky Mountains in British
      Columbia \citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

@@

      Climate forcing driving ice sheet simulations consists of
      a~present-day monthly climatology,
$\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, where temperatures are modified by offset time series,
${\Delta}T_{\text{TS}}$, and lapse-rate corrections, ${\Delta}T_{\text{LR}}$:
\begin{align}
&T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
                    + {\Delta}T_{\text{LR}}(t, x, y) \,, \\
&    P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \,.
\end{align}

      The present-day monthly climatology was computed from near-surface air
      temperature and precipitation rate fields from the NARR, averaged from
      1979 to 2000. Modern climate of the North American Cordillera is
      characterised by strong geographic variations in temperature
      seasonality, timing of the maximum annual precipitation, and daily
      temperature variability (Fig.~\ref{fig:atm}). The use of NARR is
      motivated by the need for an accurate, high-resolution precipitation
      forcing, as identified in a~previous sensitivity study
      \citep{Seguinot.etal.2014}.

@@

      Finally, lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are computed as
      a~function of ice surface elevation, $s$, using the NARR surface
      geopotential height invariant field as a~reference topography,
      $b_{\text{ref}}$:
\begin{align}
{\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{\text{ref}}],
\end{align}

      thus accounting for the evolution of ice thickness, ${h=s-b}$, on the
      one hand, and for differences between the basal topography of the ice
      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on the
      other hand. All simulations use an annual temperature lapse rate of
      $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we refer
      to different model runs by the name of the proxy record used for the
      palaeo-temperature forcing.

>>>>

      In general, the topographic complexity of the North American
      Cordillera and its effect on glacial history have inhibited the
      reconstruction of ice sheet-wide glacial advance and retreat patterns
      such as those available for the Fennoscandian and Laurentide ice
      sheets \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
      Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}. Here, we
      use a~numerical ice sheet model \citep{PISM-authors.2015}, calibrated
      against field-based evidence, to perform a~quantitative reconstruction
      of the Cordilleran ice sheet evolution through the last glacial cycle,
      and analyse some of the long-standing questions related to its
      evolution:
\begin{itemize}
  \item How much ice was locked in the Cordilleran ice sheet during the LGM?
  \item What was the scale of glaciation prior to the LGM?
  \item Which were the primary dispersal centres? Do they reflect stable or
    ephemeral configurations?
  \item How rapid was the last deglaciation? Did it include Late Glacial
    standstills or readvances?
\end{itemize}
      Although numerical ice sheet modelling has been established as
      a~useful tool to improve our understanding of the Cordilleran ice sheet
      (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991};
      \citealp{Marshall.etal.2000}), the ubiquitously mountainous topography
      of the region has presented two major challenges to its application.
      First, only recent developments in numerical ice sheet models and
      underlying scientific computing tools \citep{Bueler.Brown.2009,
      Balay.etal.2015} have allowed for high-resolution numerical modelling
      of glaciers and ice sheets on mountainous terrain over millennial time
      scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the complex
      topography of the North American Cordillera also induces strong
      geographic variations in temperature and precipitation, thus requiring
      the use of high-resolution climate forcing fields as an input to an
      ice sheet model \citep{Seguinot.etal.2014}. However, evolving climate
      conditions over the last glacial cycle are subject to considerable
      uncertainty and still lie beyond the computational reach of atmosphere
      circulation models.

@@

      The PDD computation accounts for stochastic temperature variations by
      assuming a~normal temperature distribution of standard deviation
      $\sigma$ around the expected value $T_{\mathrm{m}}$. It is expressed by
      an error-function formulation \citep{Calov.Greve.2005},
\begin{align}
&\label{eqn:calovgreve}
    {\text{PDD}} = \int_{t_1}^{t_2} \mathrm{d}t
        \left[\frac{\sigma}{\sqrt{2\pi}}
                \exp\left({-\frac{T_{\mathrm{m}}^2}{2\sigma^2}}\right)
              + \frac{T_{\mathrm{m}}}{2} \, {\text{erfc}}
                \left(-\frac{T_{\mathrm{m}}}{\sqrt{2}\sigma}\right)\right] \,,
\end{align}
      which is numerically approximated using week-long sub-intervals. In
      order to account for the effects of spatial and seasonal variations of
      temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
      from NARR daily temperature values from 1979 to 2000
      \citep{Mesinger.etal.2006}, including variability associated with the
      seasonal cycle (Fig.~\ref{fig:atm}). Degree-day factors for snow and
      ice melt are derived from mass-balance measurements on contemporary
      glaciers from the Coast Mountains and Rocky Mountains in British
      Columbia \citep[Table~\ref{tab:params};][]{Shea.etal.2009}.

@@

      Climate forcing driving ice sheet simulations consists of
      a~present-day monthly climatology, $\{T_{\mathrm{m}0},
      P_{\mathrm{m}0}\}$, where temperatures are modified by offset time
      series, ${\Delta}T_{\text{TS}}$, and lapse-rate corrections,
      ${\Delta}T_{\text{LR}}$:
\begin{align}
&T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
                    + {\Delta}T_{\text{LR}}(t, x, y) \,, \\
&    P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \,.
\end{align}
      The present-day monthly climatology was computed from near-surface air
      temperature and precipitation rate fields from the NARR, averaged from
      1979 to 2000. Modern climate of the North American Cordillera is
      characterised by strong geographic variations in temperature
      seasonality, timing of the maximum annual precipitation, and daily
      temperature variability (Fig.~\ref{fig:atm}). The use of NARR is
      motivated by the need for an accurate, high-resolution precipitation
      forcing, as identified in a~previous sensitivity study
      \citep{Seguinot.etal.2014}.

@@

      Finally, lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are computed as
      a~function of ice surface elevation, $s$, using the NARR surface
      geopotential height invariant field as a~reference topography,
      $b_{\text{ref}}$:
\begin{align}
{\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{\text{ref}}],
\end{align}
      thus accounting for the evolution of ice thickness, ${h=s-b}$, on the
      one hand, and for differences between the basal topography of the ice
      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on
      the other hand. All simulations use an annual temperature lapse rate of
      $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we refer to
      different model runs by the name of the proxy record used for the
      palaeo-temperature forcing.
